---
title: |
  Modeling Titanic Survival
author:
  - name: Qiushi Yan
    affil: a
    email: "qiushi.yann@gmail.com, website: https://qiushi.rbind.io"
affiliation:
  - num: a
    address: |
      Beijing, China
abstract: |
   This case study showcases the development of a binary logistic model to predict the possibility of survival in the loss of Titanic. I demonstrate the overall modeling process, including preprocessing, exploratory analysis, feature enginerring, model fitting, adjustment, bootstrap validation and interpretation as well as other relevant techniques such as redundancy analysis and multiple imputation for missing data. The motivation and justification behind critical statistial decisions are explained. This analysis is also made fully reproducible with R code and text provided. 
keywords: |
  logistic regression; multiple imputation; model validation
bibliography: references.bib
link-citations: true
geometry: "margin=1in"
colorlinks: yes
#appendix: appendix.tex
output:
  bookdown::pdf_book:
    base_format: rticles::tf_article
    includes:
      in_header: header.tex
---

http://www.crema-research.ch/papers/2009-03.pdf

- Do human
beings behave more in line with the selfish *homo oeconomicus*, where everybody
is out for himself or herself and possibly even puts other people’s lives in danger

https://www.insider.com/titanic-secrets-facts-2018-4#at-the-memorial-of-frederick-fleet-one-of-the-lookouts-a-prankster-left-a-pair-of-binoculars-with-a-note-reading-sorry-for-bringing-these-100-years-too-late-8


http://rpubs.com/edwardcooper/titanic1

https://www.kaggle.com/mrisdal/exploring-survival-on-the-titanic/report

https://www.kaggle.com/startupsci/titanic-data-science-solutions/comments

https://www.newscientist.com/article/dn22119-sinking-the-titanic-women-and-children-first-myth/

```{r, include = FALSE}
# global chunk options 
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.align = "center",
  echo = FALSE
)
# latex print optiosn
options(prType='latex')
# load libraries 
library(dplyr) # data wrangling
library(ggplot2) # visualization
library(rms) # modeling
library(mice) # imputation for missing data
# load raw data
titanic_raw <- readr::read_csv("data/titanic3.csv") 
# modify Hmisc::latex
mylatex <- function (...) {
    o <- capture.output(latex(...))
    # this will strip /all/ line-only comments
    o <- grep('^%', o, inv=T, value=T)
    cat(o, sep='\n')
}
# theme for ggplot2
theme_clean <- function(base_size = 11,
                      strip_text_size = 12,
                      strip_text_margin = 5,
                      subtitle_size = 13,
                      subtitle_margin = 10,
                      plot_title_size = 16,
                      plot_title_margin = 10,
                      ...) {
  ret <- ggplot2::theme_minimal(base_size = base_size, ...)
  ret$strip.text <- ggplot2::element_text(
    hjust = 0, size = strip_text_size,
    margin = ggplot2::margin(b = strip_text_margin)
  )
  ret$plot.subtitle <- ggplot2::element_text(
    hjust = 0, size = subtitle_size,
    margin = ggplot2::margin(b = subtitle_margin)
  )
  ret$plot.title <- ggplot2::element_text(
    hjust = 0, size = plot_title_size,
    margin = ggplot2::margin(b = plot_title_margin)
  )
  ret$plot.title.position <- "plot"
  ret$panel.grid <- ggplot2::element_blank()
  ret
}
ggplot2::theme_set(theme_clean())
```




# Introduction 

The sinking of RMS Titanic brought to various machine learning competitions a quintessential dataset among others, in which one major interest is to predict possibility of survival given a number of characteristics. There are several variants of this data existed on the web, the one I will be using is accessed on [Encyclopedia Titanica](https://www.encyclopedia-titanica.org/), namely [`titanic3`](http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/titanic3info.txt), courtesy of Philip Hind, with the following variables (table \@ref(data-dictionary))






```{r data-dictionary,results="asis"}
tribble(
  ~ Variable, ~ Definition, ~ Note,
  "pclass", "Ticket class", "1 = 1st, 2 = 2nd, 3 = 3rd",
  "survival", "Survival Status", "0 = No, 1 = Yes",
  "name", "Name", "",
  "sex", "Sex", "",
  "age", "Age", "In years, some infants had fractional values",
  "sibsp", "Number of Siblings/Spouses Aboard", "",
  "parch", "Number of Parents/Children Aboard", "",
  "ticket", "Ticket Number", "",
  "fare", "Passenger Fare", "in Pre-1970 British Pounds ",
  "cabin", "Cabin", "",
  "embarked", "Port of Embarkation", "Cherbourg, Queenstown or Southampton",
  "boat", "Lifeboat", "", 
  "body", "Body Identification Number", "",
  "home.dest", "Home/Destination", ""
) %>% 
  mylatex(file = "", 
          table.env = TRUE, label = "data-dictionary", size = "small",
          caption = "Data with 1309 passengers and 14 columns")

```

This data frame recorded the survival status 1309 Titanic passengers^[The data does not involve crew members, and the total number of passengers is said to be 1317]  alongside his/her gender, age, family relations on board, ticket fare, etc. There were `r (titanic_raw %>% count(survived))[["n"]][[1]]` victims and `r (titanic_raw %>% count(survived))[["n"]][[2]]` survivors in total. 

This case study has been greatly inspired by Dr. Frank Harrell's similar one in his *Regression Modeling Strategies* [-@harrell2015regression Chapter 12] book, here I attempt to propose my own idea and interpretation of model development that is as original as possible. To ensure reproducibility, all the analysis is done in R [@R-base] and RStudio with code and text made public in this [repo](https://github.com/enixam/titanic-survival). 



- Does socioeconomic advantage mean better chance of survival? If this is the case, people with higher financial means, i.e. who live in the first class are more likely to survive. Similarly,  passengers from second class will have a higher change of survival than third class people.  

- people in their prime

- Quantify predictive ability of each predictors, i.e. which predictor is most dominant in determining whether a passenger will survive 

- Find Interactions between predictors. Specifically, there are important interactions that need extra notice. For example, it has been widely studied in sociology and anthropology that human are sometimes driven by *procreation instinct* so that social norms would entail needs to protect females of reproductive age [@frey2009surviving]^[The average peak reproductive period in females is between the ages of 16 and 35.]. Therefore, we could specify and study the interaction between age and gender. Another typical interaction is between offspring and gender. *Parental investment* suggest that women on average invest more in caring for their offspring than males. In times of a disaster,  higher opportunity cost will alert females with offspring more than others, and make them seek more aggressively for changes to secure the children as well as themselves.  

- Whether the *Women and children first* policy is respected. After the collision, The captain explicitly issued an order for women and children to be saved first. ^[Though there is no international maritime law that
enforce this chivalry spirit.]

- For those who traveled alone with no family relations on the vessel, is their survival possibility greater or less? On one hand, they are more likely to be in shortage of psychological and physical support. On the other hand, 


Here is a brief summary of the following sections 

- [Exploration](#exploration), data preprocessing based on descriptive statistics and visualization, finish with a redundancy analysis  



# Exploration {#exploratory}

## Data processing and descriptive statistics

Before any analysis, we'll start by some data munging. First exclude those variables that hardly bring any insight to the possibility of survival: `name`, `ticket`^[There may be some reasons to include ticket number since their prefix could represent placement of the cabins within the ship. However, the use of such a predictor would bring excessive degrees of freedom to the model. And their poor distribution would be declared redundant by redundant analysis (later) anyway.],  `body`, `cabin`^[Because this was primarily an identification for class and most were missing.] and `home.dest`. The `boat` column is left out for another reason, because an non-missing entry in `boat` basically means survival and missing means death. For this reason "survive" and "get a life boat" is used interchangeably in the analysis. ^[More precisely, there were 9 recorded passengers who got on the lifeboat yet died before reaching Carpathia, another RMS which spearheaded the rescue of Titanic survivors. There were also 13 passengers who survived with no boat information documented, and this is most likely due to data quality issues after looking up on Encyclopedia Titanica. Even with these exceptions, whether a passenger got on a lifeboat yields perfect prediction on his/her survival. If one fits a logistic regression model on survival based on whether `boat` is missing, the apparent accuracy will be nearly 1.]. 

Next, for purposes of interpretation we will transform `fare` into today's US dollars with correction for inflation. According to discussion [here](https://www.encyclopedia-titanica.org/community/threads/ticket-prices-then-and-now.5154/), we make the transformation

$$
\frac{\text{today's US dollar}}{\text{fare in 1912}} \approx \underbrace{5}_{\text{exchange rate then}} \times \underbrace{26}_{\text{inflation index from 1912 to 2020}}
$$



At 






Finally, a nice summary of all existing variables in the data is given by the `Hmisc::describle` function. 


```{r}
# columns to include
cols <- setdiff(names(titanic_raw), 
                c("name", "ticket", "boat", 
                  "body", "cabin", "home.dest"))
# select columns
titanic <- titanic_raw[, cols] 
# 1912 pounds to us dollars
titanic$fare <- titanic$fare * 5 * 26
```


```{r, echo = FALSE, results = "asis"}
# print a summary for the data
titanic %>%
  mutate(survived = factor(survived)) %>% 
  describe() %>%
  latex(file = "", size = "small", center = "none")
```


There are several interesting patterns to notice ^[Though this may not be relevant to the model, it is still an surprising discovery that it wasn't until the late 19th century that the idea of women traveling alone gained ground. As a result, there were nearly twice as many males passengers as females on Titanic.  In fact, only 40% female passengers have no family accompanies on the ship] 

- `age` has roughly 20% missingness. On the other hand, the variable has a nice distribution with 80% known observations  falling between 14 

- Both `sibsp` and `parch` rarely have instances larger than 3, and can be readily categorized without losing too much information. 

- The distribution of `fare` is heavily right skewed, caused by the large amount of third class passengers who make do with cheap cabins, as shown in figure \@ref(fig:skew-fare). This may suggest a log transformation in the model. 



```{r skew-fare, fig.cap = "(ref:skew-fare)", fig.height = 3, fig.width = 5}
ggplot(titanic) + 
  geom_histogram(aes(fare), 
                 color = "black", fill = "midnightblue", 
                 alpha = 0.4, bins = 50) + 
  facet_wrap(~ pclass, scales = "free_x") + 
  scale_x_log10(labels = scales::label_number_si()) + 
  labs(y = NULL, x = NULL) + 
  theme(axis.text.x = element_text(size = 8), 
        strip.text = element_text(size = 10))
```

(ref:skew-fare) More than 75% of the third class passengers (700-plus in total) purchased tickets with price lower than $2000, while the median fare for second and first class is $3861. X axis is on log 10 scale


- Approximately 20% `age` is missing. This calls for a necessary imputation method as we will see later, that survival can exhibit  

```{r univariate, fig.height = 7, fig.width  = 4, fig.cap = "Summary of relationship between survival and each predictor"}
s <- summary(survived ~ age + sex + pclass + fare + 
               cut2(sibsp, 0:3) + cut2(parch, 0:3) + embarked,
    data = titanic)
plot(s, main = "" , subtitles = FALSE, cex.axis = 0.1)
```


Given this results, the last step in  


Finally, redundant analysis 

```{r, comment = ""}
redun(~ pclass  + sex + age + cut2(sibsp, 0:3) + cut2(parch, 0:3) + 
        fare + embarked, data = titanic)
```

## Data missing patterns 

There were `r sum(titanic_raw$fare == 0, na.rm = TRUE)` passengers whose `fare` is zero, all of whom males boarding in Southampton, the start of the voyage. It is suspected that some of them may be falsely included crew members, or this could be an error in data collection. I will treat these anomalies as missing entries for simplicity. 

```{r}
titanic <- titanic %>% 
  mutate(fare = na_if(fare, 0))
```


Here we use the data before excluding irrelevant columns 

```{r, echo = TRUE}
naplot(naclus(titanic), "na per var")
plot(naclus(titanic_raw))
```

There are some simple workarounds 

- complete-case analysis: That is, we delete all incomplete observations. Needless to say this will translate into a major harm on sample size since over 60% of `boat` are missing, not to mention other columns. Even if we remove `boat` and then delete rows with missing `age` we still lose over 1/5 of data. Moreover, figures in \@ref(exploratory) have shed light on the relatively strong influence of `age` on survival. Also, the deletion of incomplete observations assumes date are missing completely at random (MCAR). When it's not the case, this could severely bias estimates of coefficients [@van2018flexible]

- single imputation: 


- multiple imputation 


For demonstration purposes I will fit two decision trees to predict





## Loess regression for nonlinear pattern 

The loess is a common nonparametric regression method to study nonlinear relationship. In the case of binary response, the fitted value at $x = x_0$ is the proportion of positive cases near the neighborhood of $x_0$ ^[with varying weights according to their distance to $x_0$]. If the trend of a loess curve shows nonmonotoncity, it is reasonbale to include that nonlienarity relationship in the model, e.g., modeling the predicotr with polynomial transformation or with splines. 

figure \@ref(fig:univariate)


## Multiple imputation 




# Modeling

the choice of model. In this setting, it is obvious that we would prefer probabilistic predictions to classification with output label 0 and 1, since we are placing emphasis upon the *tendency* of survival. And the true value of our model consist not in the decision on who will survive, but in what characteristics would increase or decrease the possibility of survival. This idea has ruled out most of the black box machine learning models for classification, say, random forest, support vector machines and neural network. Not only are they not intrinsically probability oriented, it is hard to interpret main effects and interactions as everything seems to be interacted with one another. 


## Saturated model 

First and foremost, 

The limiting sample size for binary outcome would be the number of minority class, in our case `r sum(titanic$survived == 1)`. Using the 15:1 rule, that will give us some confidence spending roughly `r round(sum(titanic$survived == 1) / 15)` parameters or degrees of freedom. 


This plot is useful in identifying weather the relationship between survival status and any predictor is flat ^[A misuse of this plot would be checking nonlinearity. Even with spline transformation and large corrected $\chi^2$ there is no guarantee for nonlinearity.]. 


likely shrinkage








## Validation

There will not be another Titanic, and any model on Titanic will not be used for prediction. Therefore, the goal of model validation is primarily to provide quantify the degree of overfitting with various bias-corrected measures. 



In the award-winning solution to this legendary dataset presented by IBM Watson, they used a holdout sample to validate their model.  https://www.fharrell.com/post/split-val/

## The final model


# Discussion 


The most decisive explanation for such effect is that first-class passengers had better access to information about the
imminent danger and were aware that the lifeboats were located close to the first class cabins. Thus, their marginal effort costs to survive were lower. In contrast, most third-class passengers had no idea where the lifeboats were located (safety drills for all passengers were introduced after the Titanic disaster), and they did not know how to reach the upper decks where the lifeboats were stowed.  

 Wyn Craig Wade

> Undoubtedly, the worst barriers were the ones within the
steerage passengers themselves. Years of conditioning as
third-class citizens led a great many of them to give up hope
as soon as the crisis became evident. ..Barriers to steerage?
Yes, but of a kind less indictable to the White Star Line than
to the whole of civilization.


A more detailed explanation of some of these measures is presented in the [appendix](#measures). 




# Conclusion 






# (APPENDIX) Appendix {-} 

# Measures used in valiation {#measures}

This will be Appendix A.


# Original Computing Environment

```{r, comment = "", echo = TRUE}
sessionInfo()
```


\nocite{*}
